---
title: "Assignment_Prabisha"
author: "Prabisha Khadka"
date: "2025-05-10"
output: html_document
---


## Step 1: Installing and Loading Required Packages

```{r packages, message=FALSE}
# Install once commented since already installed
# install.packages("tidyverse")
# install.packages("GGally")
# install.packages("caret")
# install.packages("car")
# install.packages("MASS")
#install.packages("Hmisc")
#install.packages("psych")
#install.packages("ggcorrplot")
#install.packages("kableExtra")   
#install.packages("patchwork")
#install.packages("moments")
#install.packages("Metrics")
#install.packages("rsample")
#install.packages("nortest")

```

```{r}

# Load libraries for data manipulation, visualization, and regression analysis
# Tidyverse: Core data manipulation and visualization tools (ggplot2, dplyr, etc.)
library(tidyverse)

# GGally: Used for advanced plots like correlation matrix (ggpairs)
library(GGally)

# caret: Used for splitting data into training/testing (e.g. createDataPartition)
library(caret)

# car: Provides regression diagnostics and tools
library(car)

# MASS: Includes ginv() function for matrix-based estimation (manual OLS)
library(MASS)

# Hmisc: Used for correlation significance tests (e.g. rcorr())
library(Hmisc)

# ggcorrplot: For plotting correlation matrices as heatmaps
library(ggcorrplot)

# psych: Additional descriptive statistics and multivariate tools
library(psych)

# kableExtra: For formatting and styling tables in R Markdown (e.g. model output)
library(kableExtra)

# patchwork: Used to combine multiple ggplots into a single layout (e.g. 2x2 grid)
library(patchwork)

# moments: For calculating skewness and kurtosis (used in residual analysis)
library(moments)

# Metrics: Provides performance metrics like RMSE, MAE, etc.
library(Metrics)

# rsample: Used for train-test splitting using initial_split(), training(), testing()
library(rsample)

#nortest : Load the 'nortest' package to perform Anderson–Darling and other normality tests
library(nortest)

```

## Step 2: Import Dataset

```{r import-data}
# Load independent variable dataset
X <- read.csv("C:/Users/User/Downloads/assessment/independent_varaible_prabisha.csv")

# Print original names
cat("Original column names:\n")
print(colnames(X))

# Rename to full descriptive names based on assignment mapping
colnames(X) <- c("Temperature", "Pressure", "Humidity", "ExhaustVacuum")


# Print updated names
cat("Renamed columns:\n")
print(colnames(X))


```
```{r}
# Read dependent variable (EnergyOutput)

Y_df <- read.csv("C:/Users/User/Downloads/assessment/dependent_varaible_prabisha.csv", header = TRUE)

colnames(Y_df) <- "EnergyOutput"
Y <- as.numeric(Y_df$EnergyOutput)


```

```{r}
time <- 1:nrow(X)  # Create time index for time series plots

```

```{r}
# Preview dataset structure
head(X)
head(Y)
```

```{r}
# Check the dimensions of the matrix (number of rows and columns)
dim(X)

# Check the dimensions of the matrix  (number of rows and columns)
length(Y)
```
```{r}
# Generate summary
summary_stats <- summary(data.frame(X, EnergyOutput = Y))

#  Display the summary matrix 
summary_stats %>%
  kbl(
    caption = "Summary Statistics of All Variables",
    align = "c",
    digits = 3
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, border_right = TRUE)

```


# Task 1 – Exploratory Data Analysis (EDA)
You should first perform an initial exploratory data analysis, by investigating:  
• Time series plots (of input and output signal)  
• Distribution for each signal  
• Correlation and scatter plots (between different combination of input and output signals) to 
examine their dependencies




# Task 1.1 – Time series plots (of input and output signal)  


```{r}
# Convert input variables to time series
X.ts <- ts(X, start = min(time), frequency = 1)

# Also convert output variable to time series
Y.ts <- ts(Y, start = min(time), frequency = 1)

# Define colors for the input lines
colors <- c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00")

# Set layout for 2-panel display
layout(matrix(c(1, 2), nrow = 2), heights = c(4, 1))

# Plot input time series
par(mar = c(4, 5, 4, 2), xpd = TRUE)
matplot(X.ts, type = "l", lty = 1, lwd = 1.2, col = colors,
        main = "Time Series Plot of Input Variables (X)",
        xlab = "Time Index", ylab = "Input Values",
        cex.main = 1.6, cex.lab = 1.2, cex.axis = 0.9)
grid(col = "gray85", lty = "dotted")

# Plot legend
par(mar = c(0, 0, 0, 0))
plot.new()
legend("center", legend = colnames(X),
       col = colors, lty = 1, lwd = 2,
       horiz = TRUE, bty = "n", cex = 0.9)

# Output time series plot
plot(Y.ts, main = "Time Series of Energy Output (Y)",
     xlab = "Time", ylab = "Energy Output (MW)",
     col = "black", type = "l", lwd = 1.2)

```
     
     
```{r}
# ggplot version time series
X_df <- X
X_df$Time <- time
X_long <- pivot_longer(X_df, cols = -Time, names_to = "Variable", values_to = "Value")

ggplot(X_long, aes(x = Time, y = Value, color = Variable)) +
  geom_line(linewidth = 0.3) +
  facet_wrap(~Variable, ncol = 1, scales = "free_y") +
  labs(title = "Time Series Plot of Input Signals", x = "Time", y = "Input Signal") +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5))

```


# Task 1.2 – Distribution for each signal (Histogram and Density Plot) 


```{r histogram-density}
# Ensure all values are numeric
X[] <- lapply(X, function(col) as.numeric(as.character(col)))
Y <- as.numeric(Y)
```


```{r}

# Combined Distribution of All Input Variables 
all_input_values <- unlist(X)
 density_of_X_all<- density(na.omit(all_input_values))

# Combined input signal distribution
plot(density_of_X_all, main = "Density Plot of All Input Signals", col = "blue", lwd = 2)


```

```{r}
hist(na.omit(all_input_values), freq = FALSE,
     main = "Combined Histogram and Density Plot of Input Variables",
     xlab = "Combined Input Values", col = "lightblue")
lines(density_of_X_all, col = "brown", lwd = 2)
rug(jitter(na.omit(all_input_values)))
```

```{r}

# Plot for each variable
for (var in names(X)) {
  dens <- density(na.omit(X[[var]]))
  hist(X[[var]], freq = FALSE, main = paste("Histogram & Density -", var), xlab = var, col = "lightblue")
  lines(dens, lwd = 2, col = "brown")
  rug(jitter(na.omit(X[[var]])))
}
```

```{r}
# Output variable plot
density_y <- density(na.omit(Y))
hist(Y, freq = FALSE, main = "Histogram & Density - Energy Output",
     xlab = "Energy Output (MW)", col = "lightblue")
lines(density_y, lwd = 2, col = "brown")
rug(jitter(na.omit(Y)))
```

# Task 1.3 –  Correlation and scatter plots


```{r correlation-scatter}
# Combine input and output
data_cor <- data.frame(X, EnergyOutput = Y)
data_cor <- na.omit(data_cor)

cor_test <- rcorr(as.matrix(data_cor))


```

```{r}
# 1. ggpairs: correlation + scatter + density
GGally::ggpairs(
  data_cor,
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
  upper = list(continuous = wrap("cor", size = 3)),
  diag  = list(continuous = wrap("densityDiag")),
  title = "Correlation and Scatter Plot Matrix"
)
```


```{r}
# Heatmap of correlations
corr_mat <- round(cor(data_cor), 2)
ggcorrplot::ggcorrplot(corr_mat, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3, colors = c("red", "white", "blue"),
           title = "Correlation Heatmap", ggtheme = theme_minimal())
```
```{r}
#  Base R scatter plots
par(mfrow = c(2, 2))
plot(X$Temperature, Y, main = "Temperature vs Energy Output", xlab = "Temperature", ylab = "Energy Output", col = "blue", pch = 20)
plot(X$ExhaustVacuum, Y, main = "ExhaustVacuum vs Energy Output", xlab = "ExhaustVacuum", ylab = "Energy Output", col = "darkgreen", pch = 20)
plot(X$Pressure, Y, main = "Pressure vs Energy Output", xlab = "Pressure", ylab = "Energy Output", col = "purple", pch = 20)
plot(X$Humidity, Y, main = "Humidity vs Energy Output", xlab = "Humidity", ylab = "Energy Output", col = "orange", pch = 20)
```

```{r}
# Remove row names
rownames(cor_test$r) <- NULL

# Correlation Coefficients Table
round(cor_test$r, 3) %>%
  kbl(
    caption = "Correlation Coefficients (r) Between Variables",
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white")


```

```{r}
# Fit linear model and compute VIF
vif_model <- lm(EnergyOutput ~ ., data = data_cor)
car::vif(vif_model)
```

```{r}
#  Correlation Matrix and Heatmap
combined_data <- data.frame(X, EnergyOutput = Y)
cor_matrix <- cor(combined_data)
ggcorrplot::ggcorrplot(cor_matrix, lab = TRUE, title = "Correlation Heatmap")

```

# Task 2 - Regression – modelling the relationship between gene expression  
We would like to determine a suitable mathematical model in explaining the relationship between the output Net hourly electrical energy (y) = x2 with other input x2: Net hourly electrical energy output (EP) in MW (dependent variable), x1: Temperature (T) – Ambient temperature 
(°C), x3: Ambient Pressure (AP) – Atmospheric pressure (millibar), x4: Relative Humidity (RH) – Humidity level (%), x5: Exhaust Vacuum (V) – Vacuum collected from the steam turbine (cm Hg) that ‘regulate’ its expression, which we assume can be described by a polynomial 
regression model. Below are 5 candidate nonlinear polynomial regression models, and only one of them can ‘truly’ describe such a relationship?
The objective is to identify this ‘true’ model from those candidate models following Tasks 2.1 – 2.6. To accomplish these objectives, understanding the interconnection between different variable is crucial, which can be achieved through modeling and analyzing provided data. 
Data sets: Provided in https://schoolworkspro.com.

# Task 2.1: Estimation of Model Parameters using Least Squares


```{r}
# Combine input (X) and output (Y)
data_df <- data.frame(
  x1 = X$Temperature,
  x3 = X$Pressure,
  x4 = X$Humidity,
  x5 = X$ExhaustVacuum,
  x2 = Y
)
```

```{r}

# Define response variable
y_response <- data_df$x2
head(y_response)
```

```{r}
#  Define model matrices with comments
# Model 1: y = θ₁·x₄ + θ₂·x₃² + θ_bias
X_model1 <- cbind(1, data_df$x4, data_df$x3^2)
colnames(X_model1) <- c("Intercept", "x4", "x3_sq")
head(X_model1)
```

```{r}
# Model 2: y = θ₁·x₄ + θ₂·x₃² + θ₃·x₅ + θ_bias
X_model2 <- cbind(1, data_df$x4, data_df$x3^2, data_df$x5)
colnames(X_model2) <- c("Intercept", "x4", "x3_sq", "x5")
head(X_model2)
```

```{r}
# Model 3: y = θ₁·x₃ + θ₂·x₄ + θ₃·x₅³ (no intercept)
X_model3 <- cbind(data_df$x3, data_df$x4, data_df$x5^3)
colnames(X_model3) <- c("x3", "x4", "x5_cubed")
head(X_model3)
```

```{r}
# Model 4: y = θ₁·x₄ + θ₂·x₃² + θ₃·x₅³ + θ_bias
X_model4 <- cbind(1, data_df$x4, data_df$x3^2, data_df$x5^3)
colnames(X_model4) <- c("Intercept", "x4", "x3_sq", "x5_cubed")
head(X_model4)
```

```{r}
# Model 5: y = θ₁·x₄ + θ₂·x₁² + θ₃·x₃² + θ_bias
X_model5 <- cbind(1, data_df$x4, data_df$x1^2, data_df$x3^2)
colnames(X_model5) <- c("Intercept", "x4", "x1_sq", "x3_sq")
head(X_model5)
```

```{r}
# Step 4: Estimate theta (OLS or ginv)
Model1_thetahat <- solve(t(X_model1) %*% X_model1) %*% t(X_model1) %*% y_response
Model1_thetahat
```

```{r}
Model2_thetahat <- solve(t(X_model2) %*% X_model2) %*% t(X_model2) %*% y_response
Model2_thetahat
```

```{r}
Model3_thetahat <- solve(t(X_model3) %*% X_model3) %*% t(X_model3) %*% y_response
Model3_thetahat
```

```{r}
Model4_thetahat <- ginv(t(X_model4) %*% X_model4) %*% t(X_model4) %*% y_response
Model4_thetahat
```

```{r}
Model5_thetahat <- solve(t(X_model5) %*% X_model5) %*% t(X_model5) %*% y_response
Model5_thetahat
```

```{r}
# Create the summary table
theta_table <- data.frame(
  Model  = paste0("Model", 1:5),
  Bias   = c(Model1_thetahat[1], Model2_thetahat[1], NA, Model4_thetahat[1], Model5_thetahat[1]),
  Theta1 = c(Model1_thetahat[2], Model2_thetahat[2], Model3_thetahat[1], Model4_thetahat[2], Model5_thetahat[2]),
  Theta2 = c(Model1_thetahat[3], Model2_thetahat[3], Model3_thetahat[2], Model4_thetahat[3], Model5_thetahat[3]),
  Theta3 = c(NA, Model2_thetahat[4], Model3_thetahat[3], Model4_thetahat[4], Model5_thetahat[4])
)

# Display thetahat table (styled)
theta_table %>%
  mutate(across(where(is.numeric), round, 5)) %>%
  kbl(
    caption = "Estimated Thetahat Coefficients for All Models",
    col.names = c("Model", "Bias", "Theta 1", "Theta 2", "Theta 3"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, border_right = TRUE) %>%
  column_spec(2:5, width = "3.5cm")
```


```{r}
# Matrix diagnostics – Rank and Condition Number
for (i in 1:5) {
  Xmat <- get(paste0("X_model", i))
  cat(paste0("Model ", i, " - Rank: ", qr(Xmat)$rank,
             ", Condition Number: ", round(kappa(Xmat), 2), "\n"))
}
```

# Task 2.2: Compute Residual Sum of Squares (RSS)


```{r}
#  Residual Mean Check (Should ≈ 0)
for (i in 1:5) {
  Xmat <- get(paste0("X_model", i))
  theta <- get(paste0("Model", i, "_thetahat"))
  y_pred <- Xmat %*% theta
  resids <- y_response - y_pred
  cat(paste0("Model ", i, " - Mean Residual: ", round(mean(resids), 5), "\n"))
}

```

```{r}
# Save residuals for further anlaysis
residuals_all_models <- list(
  model1 = y_response - X_model1 %*% Model1_thetahat,
  model2 = y_response - X_model2 %*% Model2_thetahat,
  model3 = y_response - X_model3 %*% Model3_thetahat,
  model4 = y_response - X_model4 %*% Model4_thetahat,
  model5 = y_response - X_model5 %*% Model5_thetahat
)
```

```{r}
# Step 1: Compute RSS for each model
rss_values <- sapply(residuals_all_models, function(resid) sum(resid^2))

# Step 2: Create rss_df to store Model labels and values
rss_df <- data.frame(
  Model = paste0("Model ", 1:5),
  RSS = round(rss_values, 2)
)

# Step 3: Display RSS table
rownames(rss_df) <- NULL

library(knitr)
library(kableExtra)

rss_df %>%
  kbl(
    caption = "Task 2.2: Residual Sum of Squares (RSS) for All Models",
    col.names = c("Model", "RSS"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, width = "6cm", border_right = TRUE) %>%
  column_spec(2, width = "8cm")
```


# Task 2.3: Compute Log-Likelihood



```{r}
n <- length(y_response)  # Number of samples

# Calculate Variance and Log-Likelihood
variance_values <- rss_values / (n - 1)

loglik_values <- - (n / 2) * log(2 * pi) -
                 (n / 2) * log(variance_values) -
                 (1 / (2 * variance_values)) * rss_values

# Combine into a table
loglik_df <- data.frame(
  Model = paste0("Model ", 1:5),
  Variance = round(variance_values, 5),
  LogLikelihood = round(loglik_values, 2)
)

# Display the table of Variance and Log-Likelihood for All Models
loglik_df %>%
  kbl(
    caption = "Variance and Log-Likelihood for All Models",
    col.names = c("Model", "Variance", "Log-Likelihood"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2e4a62", color = "white") %>%
  column_spec(1, bold = TRUE, width = "6cm", border_right = TRUE) %>%
  column_spec(2, width = "6cm") %>%
  column_spec(3, width = "6cm")

```


# Task 2.4 – AIC and BIC for All Models


```{r}
#  Sample size
n <- length(y_response)

#  Number of parameters (k) for each model
k_values <- c(
  length(Model1_thetahat),
  length(Model2_thetahat),
  length(Model3_thetahat),
  length(Model4_thetahat),
  length(Model5_thetahat)
)

# AIC and BIC calculations
aic_values <- 2 * k_values - 2 * loglik_values
bic_values <- k_values * log(n) - 2 * loglik_values

# Create result data frame
aicbic_df <- data.frame(
  Model = paste0("Model ", 1:5),
  Parameters = k_values,
  AIC = round(aic_values, 2),
  BIC = round(bic_values, 2)
)

# Print AIC & BIC values
cat("AIC and BIC for all models:\n")
for (i in 1:5) {
  cat(paste0("Model ", i,
             " → Parameters: ", k_values[i],
             " | AIC: ", round(aic_values[i], 2),
             " | BIC: ", round(bic_values[i], 2), "\n"))
}
```


```{r}
# Remove row names to avoid 'model1', 'model2', etc.
rownames(aicbic_df) <- NULL

# Display table of AIC and BIC for All Models
aicbic_df %>%
  kbl(
    caption = "AIC and BIC for All Models",
    col.names = c("Model", "Parameters (k)", "AIC", "BIC"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, width = "6cm", border_right = TRUE) %>%
  column_spec(2:4, width = "5cm")
  
```



# Task 2.5 – Q-Q Plot of Residuals to Assess Normality

```{r}
# reusable Q–Q plot function 
plot_qq <- function(residuals, model_label) {
  df <- data.frame(residuals = residuals)
  
  ggplot(df, aes(sample = residuals)) +
    stat_qq(color = "#1f78b4", size = 1.5) +  # deep blue points
    stat_qq_line(color = "#003366", linetype = "dashed", linewidth = 1) +  # navy reference line
    labs(title = paste("Q–Q Plot for", model_label),
         x = "Theoretical Quantiles",
         y = "Sample Quantiles") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#003366"),
      axis.title = element_text(face = "bold", size = 12, color = "#003366"),
      axis.text = element_text(color = "#2c3e50"),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()
    )
}

```


```{r}

# Model 1
plot_qq(residuals_all_models$model1, "Model 1")
```
```{r}
# Model 2
plot_qq(residuals_all_models$model2, "Model 2")
```

```{r}
# Model 3
plot_qq(residuals_all_models$model3, "Model 3")
```


```{r}
# Model 4
plot_qq(residuals_all_models$model4, "Model 4")
```

```{r}
# Model 5
plot_qq(residuals_all_models$model5, "Model 5")
```

```{r}

# Define the plotting function 
plot_histogram <- function(residuals, model_label) {
  df <- data.frame(residuals = residuals)
  
  ggplot(df, aes(x = residuals)) +
    geom_histogram(aes(y = after_stat(density)), fill = "#1f78b4", alpha = 0.7, bins = 30) +
    geom_density(color = "#e74c3c", linewidth = 1.2, linetype = "dashed") +
    labs(title = model_label, x = "Residuals", y = "Density") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5, color = "#003366"),
      axis.title = element_text(face = "bold", size = 12, color = "#003366"),
      axis.text = element_text(color = "#2c3e50")
    )
}

# Step 1: Generate plots in a loop
histogram_list <- list()

for (i in 1:5) {
  model_name <- paste("Model", i)
  resids <- residuals_all_models[[paste0("model", i)]]
  histogram_list[[i]] <- plot_histogram(resids, model_name)
}

# Step 2: Arrange plots into a 3+2 grid layout
(histogram_list[[1]] + histogram_list[[2]] + histogram_list[[3]]) /
(histogram_list[[4]] + histogram_list[[5]] + plot_spacer()) +
  plot_annotation(
    title = "Residual Histograms with Density for All Models",
    theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
  )
```
```{r}
cat("Normality Test Results:\n")
for (i in 1:5) {
  resids <- residuals_all_models[[paste0("model", i)]]
  sample_resids <- if (length(resids) > 5000) sample(resids, 5000) else resids
  
  shapiro <- shapiro.test(sample_resids)
  ad <- ad.test(resids)
  
  cat(paste0("Model ", i,
             " | Shapiro p = ", round(shapiro$p.value, 4),
             " | A–D stat = ", round(ad$statistic, 4),
             ", p = ", round(ad$p.value, 4), "\n"))
}

```

```{r}
# Calculate normality test results
shapiro_p <- numeric(5)
ad_stat <- numeric(5)
ad_p <- numeric(5)


norm_test_summary <- data.frame(
  Model = paste0("Model ", 1:5),
  Shapiro_p = shapiro_p,
  AD_stat = ad_stat,
  AD_p = ad_p
)
```

```{r}
# Create residual summary for all models
res_summary <- data.frame(
  Model = paste0("Model ", 1:5),
  Mean = sapply(residuals_all_models, function(r) round(mean(r), 4)),
  SD = sapply(residuals_all_models, function(r) round(sd(r), 4)),
  Skewness = sapply(residuals_all_models, function(r) round(skewness(r), 4)),
  Kurtosis = sapply(residuals_all_models, function(r) round(kurtosis(r), 4))
)
```

```{r}
# Remove default row names
rownames(norm_test_summary) <- NULL

# Display table with consistent styling
norm_test_summary %>%
  kbl(
    caption = "Task 2.5: Normality Test Summary (Shapiro–Wilk and Anderson–Darling)",
    col.names = c("Model", "Shapiro–Wilk p-value", "A–D Statistic", "A–D p-value"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, width = "5cm", border_right = TRUE) %>%
  column_spec(2:4, width = "4cm")

```
```{r}
# Remove row names to avoid `model1`, `model2` in output
rownames(res_summary) <- NULL

# Display with same clean styling as AIC/BIC table
res_summary %>%
  kbl(
    caption = "Task 2.5: Residual Summary Statistics for All Models",
    col.names = c("Model", "Mean", "Standard Deviation", "Skewness", "Kurtosis"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, width = "5cm", border_right = TRUE) %>%
  column_spec(2:5, width = "4cm")
```


# Task 2.6 – Select Best Regression Model Based on AIC, BIC & Residuals

```{r}
#  Combine all key metrics
model_summary <- data.frame(
  Model = paste0("Model ", 1:5),
  Parameters = k_values,
  RSS = round(rss_values, 2),
  Variance = round(variance_values, 5),
  AIC = round(aic_values, 2),
  BIC = round(bic_values, 2),
  SD = sapply(residuals_all_models, function(r) round(sd(r), 4)),
  Skewness = sapply(residuals_all_models, function(r) round(skewness(r), 4)),
  Kurtosis = sapply(residuals_all_models, function(r) round(kurtosis(r), 4))
)
```

```{r}

rownames(model_summary) <- NULL

# Display the final summary table
model_summary %>%
  kbl(
    caption = "Task 2.6: Final Model Comparison Summary",
    col.names = c("Model", "Parameters", "RSS", "Variance", "AIC", "BIC", "SD", "Skewness", "Kurtosis"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, width = "5.5cm", border_right = TRUE)
```
```{r}
#  Determine best model based on AIC and BIC
best_model_index <- which.min(aic_values)
chosen_model <- paste0("Model ", best_model_index)
```

```{r}
# Print a justification
cat("Selected Best Model:", chosen_model, "\n\n")

cat("Justification:\n")
cat("- It has the lowest AIC and BIC among all models.\n")
cat("- It maintains good residual properties (mean ~0, moderate skewness/kurtosis).\n")
cat("- Predictors are interpretable: Temperature², Pressure², and Humidity align with physical domain knowledge.\n")
cat("- Q–Q plots and histograms support approximate residual normality.\n")
```

```{r}
#  AIC comparison bar plot
barplot(
  aic_values,
  names.arg = paste("Model", 1:5),
  col = "#1f78b4",
  main = "AIC Comparison Across Models",
  ylab = "AIC Score",
  ylim = c(min(aic_values) * 0.99, max(aic_values) * 1.02)
)
```

```{r}
# Print the chosen model based on AIC and BIC
best_model_index <- which.min(aic_values)
cat("Selected Best Model:", paste0("Model ", best_model_index),
    "\nJustification: Lowest AIC and BIC, good residual distribution, and interpretable terms.\n")
```

# Task 2.7 – Train/Test Split, Prediction, CI, Metrics, and Plot


```{r}
# Step 1: Prepare full data for Model 5 (x4, x1², x3²)
data_full <- data.frame(
  x4 = X$Humidity,
  x1_sq = X$Temperature^2,
  x3_sq = X$Pressure^2,
  y = Y
)

# Step 2: Train-test split (70% training)
set.seed(123)
split_data <- initial_split(data_full, prop = 0.7)
train_data <- training(split_data)
test_data  <- testing(split_data)

# Step 3: Fit Model 5 on training set
model5_train <- lm(y ~ x4 + x1_sq + x3_sq, data = train_data)

# Step 4: Predict on test set with 95% confidence interval
pred_test <- predict(model5_train, newdata = test_data, interval = "confidence", level = 0.95)

# Step 5: Prepare results dataframe
results <- data.frame(
  Index = 1:nrow(test_data),
  Observed = test_data$y,
  Predicted = pred_test[, "fit"],
  Lower_CI = pred_test[, "lwr"],
  Upper_CI = pred_test[, "upr"]
)

# Step 6: Plot predictions with CI
ggplot(results, aes(x = Index)) +
  geom_point(aes(y = Observed), color = "#1f78b4", size = 1.2, alpha = 0.5) +
  geom_line(aes(y = Predicted), color = "#e31a1c", linewidth = 1.2) +
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI), fill = "gray70", alpha = 0.3) +
  labs(
    title = " Model 5 Prediction with 95% Confidence Interval",
    x = "Test Sample Index", y = "Energy Output"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", size = 12, color = "#003366")
  )
```

```{r}
# Calculate model performance metrics (RMSE, MAE, R²)
test_resids <- test_data$y - pred_test[, "fit"]

rmse_val <- Metrics::rmse(test_data$y, pred_test[, "fit"])
mae_val  <- Metrics::mae(test_data$y, pred_test[, "fit"])
r2_val   <- 1 - sum(test_resids^2) / sum((test_data$y - mean(test_data$y))^2)

```

```{r}
# Create summary table of performance metrics
metrics_df <- data.frame(
  Metric = c("RMSE", "MAE", "R²"),
  Value = c(round(rmse_val, 4), round(mae_val, 4), round(r2_val, 4))
)

metrics_df %>%
  kbl(
    caption = "Model 5 Performance Metrics on Test Data",
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white") %>%
  column_spec(1, bold = TRUE, width = "6cm", border_right = TRUE)

```
```{r}
# Step 10: Visualize test residuals
ggplot(data.frame(test_resids), aes(x = test_resids)) +
  geom_histogram(bins = 30, fill = "#1f78b4", alpha = 0.7) +
  geom_density(color = "red", linewidth = 1.2, linetype = "dashed") +
  labs(
    title = "Distribution of Test Residuals (Model 5)",
    x = "Residuals", y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", color = "#003366")
  )
```

```{r}
#  Actual vs. Predicted Plot
ggplot(results, aes(x = Observed, y = Predicted)) +
  geom_point(color = "#1f78b4", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title = "Actual vs Predicted Energy Output (Test Data)",
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", color = "#003366")
  )

```
```{r}
# Density plot of training Y 
ggplot(data.frame(y_train = train_data$y), aes(x = y_train)) +
  geom_density(fill = "#1f78b4", alpha = 0.6) +
  geom_vline(xintercept = mean(train_data$y), color = "red", linetype = "dashed") +
  labs(
    title = "Density Plot of Training Energy Output (Y)",
    x = "Energy Output", y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", color = "#003366")
  )
```
```{r}
# Using training data from split
X_manual <- cbind(
  1,
  train_data$x4,
  train_data$x1_sq,
  train_data$x3_sq
)

Y_manual <- as.matrix(train_data$y)
thetaHat_training <- ginv(t(X_manual) %*% X_manual) %*% t(X_manual) %*% Y_manual

# Using MASS::ginv

thetaHat_training <- ginv(t(X_manual) %*% X_manual) %*% t(X_manual) %*% Y_manual

theta_labels <- c("θ₀ (bias)", "θ₁ (Humidity)", "θ₂ (Temperature²)", "θ₃ (Pressure²)")
theta_values <- as.numeric(thetaHat_training)

data.frame(Parameter = theta_labels, Estimate = round(theta_values, 6)) %>%
  kbl(caption = "Manually Estimated Model Parameters (Training Data)") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white")

```

```{r}
round(coef(model5_train), 6)
```

# Task 3 –  Approximate Bayesian Computation (ABC)  
Using ‘rejection ABC’ method to compute the posterior distributions of the ‘selected’  regression model parameters in Task 2.  
1) You only need to compute 2 parameter posterior distributions -- the 2 parameters withlargest absolute value in your least squares estimation (Task 2.1) of the selected model. Fix all the other parameters in your model as constant, by using the estimated values from Task 2.1.  
2) Use a Uniform distribution as prior, around the estimated parameter values for those 2 parameters (from Task 2.1). You will need to determine the range of the prior distribution. 
3) Draw samples from the above Uniform prior and perform rejection ABC for those 2 parameters.
4) Plot the joint and marginal posterior distribution for those 2 parameters.  
5) Explain your results. 


```{r}
# Fixed values from Task 2.1
theta_fixed <- c(
  intercept = 251.6717,
  humidity  = 0.2272,
  temp_sq   = -0.0365,
  press_sq  = -0.0037
)

# Define Uniform priors for θ₀ and θ₁ (±20%)
prior_theta0 <- c(0.8 * theta_fixed["intercept"], 1.2 * theta_fixed["intercept"])
prior_theta1 <- c(0.8 * theta_fixed["humidity"], 1.2 * theta_fixed["humidity"])

# Design matrix for Model 5
X_model5 <- cbind(
  1,
  X$Humidity,
  X$Temperature^2,
  X$Pressure^2
)

#  Baseline Ŷ and RSS
Y_hat_base <- X_model5 %*% theta_fixed
RSS_base <- sum((Y - Y_hat_base)^2)

# Rejection ABC Setup
set.seed(123)
n_samples <- 5000
epsilon <- RSS_base * 1.2

accepted_theta0 <- c()
accepted_theta1 <- c()

# Rejection Sampling Loop
for (i in 1:n_samples) {
  theta0_sample <- runif(1, prior_theta0[1], prior_theta0[2])
  theta1_sample <- runif(1, prior_theta1[1], prior_theta1[2])
  
  theta_proposed <- c(
    theta0_sample,
    theta1_sample,
    theta_fixed["temp_sq"],
    theta_fixed["press_sq"]
  )
  
  Y_hat_sim <- X_model5 %*% theta_proposed
  RSS_sim <- sum((Y - Y_hat_sim)^2)
  
  if (RSS_sim < epsilon) {
    accepted_theta0 <- c(accepted_theta0, theta0_sample)
    accepted_theta1 <- c(accepted_theta1, theta1_sample)
  }
}

cat("ABC complete — Accepted samples:", length(accepted_theta0), "out of", n_samples, "\n")
```

```{r}
# Combine accepted samples from rejection ABC into a data frame
posterior_df <- data.frame(
  theta0 = accepted_theta0,
  theta1 = accepted_theta1
)
```
```{r}
# Joint Posterior Plot
ggplot(posterior_df, aes(x = theta0, y = theta1)) +
  geom_point(alpha = 0.5, color = "#1f78b4") +
  labs(
    title = "Joint Posterior Distribution through ABC",
    x = expression(theta[0] ~ "(Intercept)"),
    y = expression(theta[1] ~ "(Humidity Coefficient)")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", color = "#003366")
  )
```


```{r}
# Marginal Posterior of θ₀ (Intercept)
ggplot(posterior_df, aes(x = theta0)) +
  geom_histogram(bins = 30, fill = "#1f78b4", alpha = 0.7) +
  geom_vline(xintercept = theta_fixed["intercept"], color = "black", linetype = "dashed", linewidth = 1) +
  labs(
    title = expression("Marginal Posterior of " ~ theta[0]),
    x = expression(theta[0]), y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", color = "#003366")
  )

```

```{r}

#  Marginal Posterior of θ₁ (Humidity)
ggplot(posterior_df, aes(x = theta1)) +
  geom_histogram(bins = 30, fill = "#e31a1c", alpha = 0.7) +
  geom_vline(xintercept = theta_fixed["humidity"], color = "black", linetype = "dashed", linewidth = 1) +
  labs(
    title = expression("Marginal Posterior of " ~ theta[1]),
    x = expression(theta[1]), y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, color = "#003366"),
    axis.title = element_text(face = "bold", color = "#003366")
  )

```
```{r}
# Create posterior summary table
posterior_summary <- data.frame(
  Parameter = c("θ₀ (Intercept)", "θ₁ (Humidity)"),
  Mean = c(mean(accepted_theta0), mean(accepted_theta1)),
  SD   = c(sd(accepted_theta0), sd(accepted_theta1))
)

posterior_summary %>%
  kbl(
    caption = "Posterior Summary (ABC Accepted Parameters)",
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#2c3e50", color = "white")
```






